A pipeline to analyse time-course gene expression data

### 2019-08-28 scripts/manuscript.Rmd
Use of newer moanin functions/names

**Use with `moanin`** `moanin` provides functionality for performing both of
these types of approaches, though our focus is on the global approach. In both
situations, we first need to set up a object (a `splines_model` object) to
hold the meta data, as well as information for fitting the spline model.

We start by creating the `splines_model` object using the
`create_splines_model` function.

```{r}
splines_model = create_splines_model(meta, degrees_of_freedom=6)
```

The `splines_model` object contains a number of metadata and options used
throughout this analysis: the condition and timepoints of each samples, the
formula object used or the basis or the degrees of freedom of the model. The
metadata data.frame object should contain at least two columns: one named
`Group`, containing the  treatement effect, and a second one named `Timepoint`
containing the timepoint information.

@@

```{r make_timepoints_contarts}
contrasts = create_timepoints_contrasts("M", "VL", splines_model)
```

@@

```{r weekly_differential_analysis}
# Define contrast	
weekly_de_analysis = DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)
```
**EAP: Sort of strange to have argument `spines_model` for function that doesn't use splines. Perhaps there is a generic name that could be used for the argument**

This results in a table of the results, where each row corresponds to a gene
and the columns correspond to the p-value (`pval`), log-fold change (`lfc`)
and adjusted p-value (`qval`); the order of the genes in the table is the same
as the input `data`. Here we show the results for the first two timepoints and
the first ten genes:

@@

**EAP: For the output of this, the column names are in format `M.0-VL.0-pval`, for example. This may be hard to parse algorithmically down the road because you are using the same character for the contrast (`-`) as the separator for the `pval`. So for example, if you wanted to split the contrast name from the `pval` or `lfc` part, it wouldn't be very easy. Maybe underscore `_` would be better separator?**

@@

```{r echo=FALSE, results="hide"}
contrasts = create_timepoints_contrasts("M", "K", splines_model)
K_weekly_de_analysis =  DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts("M", "C", splines_model)
C_weekly_de_analysis =  DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts("M", "VH", splines_model)
VH_weekly_de_analysis =  DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)

@@

```{r showMissingDEGenes, echo=FALSE}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
plot_splines_data(data, splines_model, subset_data=exampleGenes,
		  colors=ann_colors$Group)
```

@@

# The function takes the data (data.frame or named matrix), the meta data
# (data.frame containing a timepoint and group column, the first corresponding
# to the time-course information, the latter corresponding to the
# treatment).
pvalues = DE_timecourse(
    data, splines_model, timecourse_contrasts,
    use_voom_weights=FALSE)
qvalues = apply(pvalues, 2, p.adjust)

@@

```{r}
log_fold_change_timepoints = estimate_log_fold_change(
    data, splines_model, timecourse_contrasts,  method="timely")
```

@@

```{r}
log_fold_change_timecourse = estimate_log_fold_change(
    data, splines_model, timecourse_contrasts,  method="timecourse")

log_fold_change_sum = estimate_log_fold_change(
    data, splines_model, timecourse_contrasts,  method="sum")

log_fold_change_max = estimate_log_fold_change(
    data, splines_model, timecourse_contrasts, method="max")

log_fold_change_min = estimate_log_fold_change(
    data, splines_model, timecourse_contrasts, method="min")
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:10])
plot_splines_data(data[top_DE_genes_pval, ], splines_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_timecourse),
	 decreasing=TRUE)[1:10])
plot_splines_data(data[top_DE_genes_lfc, ], splines_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))

@@

```{r clustering}
# First fit the kmeans clusters
kmeans_clusters = splines_kmeans(
    y, splines_model, n_clusters=8,
    random_seed=42,
    n_init=20)
```

@@

```{r fig.height=6, fig.width=6}
plot_splines_data(
    kmeans_clusters$centroids, splines_model,
    colors=ann_colors$Group,
    smooth=TRUE)
```

@@

i = 1
for(n_cluster in all_possible_n_clusters){
    clustering_results = splines_kmeans(
	y, splines_model,
	n_clusters=n_cluster, random_seed=42,
	n_init=10)
    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
    all_clustering[[i]] = clustering_results$clusters
    i = i + 1
}
```

>>>>

**Use with `moanin`** `moanin` provides functionality for performing both of
these types of approaches, though our focus is on the global approach. In both
situations, we first need to set up a object (a `moanin_model` object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, …)

We start by creating the `moanin_model` object using the
`create_moanin_model` function.

```{r}
moanin_model = create_moanin_model(meta, degrees_of_freedom=6)
```

The `moanin_model` object contains a number of metadata and options used
throughout this analysis: the condition and timepoints of each samples, the
formula object used or the basis or the degrees of freedom of the model. The
metadata data.frame object should contain at least two columns: one named
`Group`, containing the  treatement effect, and a second one named `Timepoint`
containing the timepoint information.

@@

```{r make_timepoints_contarts}
contrasts = create_timepoints_contrasts("M", "VL", moanin_model)
```

@@

```{r weekly_differential_analysis}
# Define contrast	
weekly_de_analysis = DE_timepoints(
     data, moanin_model, contrasts,
     use_voom_weights=FALSE)
```
This results in a table of the results, where each row corresponds to a gene
and the columns correspond to the p-value (`pval`), log-fold change (`lfc`)
and adjusted p-value (`qval`); the order of the genes in the table is the same
as the input `data`. Here we show the results for the first two timepoints and
the first ten genes:

@@

@@

```{r echo=FALSE, results="hide"}
contrasts = create_timepoints_contrasts("M", "K", moanin_model)
K_weekly_de_analysis =  DE_timepoints(
     data, moanin_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts("M", "C", moanin_model)
C_weekly_de_analysis =  DE_timepoints(
     data, moanin_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts("M", "VH", moanin_model)
VH_weekly_de_analysis =  DE_timepoints(
     data, moanin_model, contrasts,
     use_voom_weights=FALSE)

@@

```{r showMissingDEGenes, echo=FALSE}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
plot_splines_data(data, moanin_model, subset_data=exampleGenes,
		  colors=ann_colors$Group)
```

@@

# The function takes the data (data.frame or named matrix), the meta data
# (data.frame containing a timepoint and group column, the first corresponding
# to the time-course information, the latter corresponding to the
# treatment).
pvalues = DE_timecourse(
    data, moanin_model, timecourse_contrasts,
    use_voom_weights=FALSE)
qvalues = apply(pvalues, 2, p.adjust)

@@

```{r}
log_fold_change_timepoints = estimate_log_fold_change(
    data, moanin_model, timecourse_contrasts,  method="timely")
```

@@

To compute the log-fold change, `moanin` provides a function taking as
argument the data, the `moanin_model` object, the contrasts to evaluate, as
well as the method to use to estimate the log-fold change.

```{r}
log_fold_change_timecourse = estimate_log_fold_change(
    data, moanin_model, timecourse_contrasts,  method="timecourse")

log_fold_change_sum = estimate_log_fold_change(
    data, moanin_model, timecourse_contrasts,  method="sum")

log_fold_change_max = estimate_log_fold_change(
    data, moanin_model, timecourse_contrasts, method="max")

log_fold_change_min = estimate_log_fold_change(
    data, moanin_model, timecourse_contrasts, method="min")
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:10])
plot_splines_data(data[top_DE_genes_pval, ], moanin_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_timecourse),
	 decreasing=TRUE)[1:10])
plot_splines_data(data[top_DE_genes_lfc, ], moanin_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))

@@

```{r clustering}
# First fit the kmeans clusters
kmeans_clusters = splines_kmeans(
    y, moanin_model, n_clusters=8,
    random_seed=42,
    n_init=20)
```

@@

```{r fig.height=6, fig.width=6}
plot_splines_data(
    kmeans_clusters$centroids, moanin_model,
    colors=ann_colors$Group,
    smooth=TRUE)
```

@@

i = 1
for(n_cluster in all_possible_n_clusters){
    clustering_results = splines_kmeans(
	y, moanin_model,
	n_clusters=n_cluster, random_seed=42,
	n_init=10)
    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
    all_clustering[[i]] = clustering_results$clusters
    i = i + 1
}
```
